% ============================================================
% MCCA Paper — Appendix: v1.1 Upgrade Plan
% ============================================================

\appendix

\section{v1.1 Upgrade Plan}
\label{app:v11}

This appendix describes the planned v1.1 expansion of the empirical
validation suite and threat model registry.

\subsection{Threshold calibration}

The following metrics are observational at v0.1 and become gating in
v1.1 once empirical baselines are established:

\begin{itemize}
  \item \textbf{Adversarial Detection Rate (HAR-SPEC-005, S3):}
        ADR threshold raised from $\geq 0.60$ to a value calibrated
        against the v0.1 empirical ADR distribution. Target: $\geq 0.80$.
  \item \textbf{Deviation Score (HAR-SPEC-012, T3):}
        DevS threshold set from v0.1 observational distribution.
        Becomes a soft gate in v1.1.
  \item \textbf{Stability Score (HAR-SPEC-012, T3):}
        SS threshold set from v0.1 re-run data ($k \geq 3$).
        Distinguishes persistent deviations from one-off artifacts.
\end{itemize}

\subsection{New threat coverage}

The two \texttt{none}-coverage threats deferred from v0.1:

\begin{itemize}
  \item \textbf{TM-009 specification\_gaming:} Requires a simulator
        scenario (SIM-0XX) and harness probe (HAR-SPEC-009) testing
        whether the model satisfies literal governance constraints while
        violating intended semantics. Detection depends on semantic
        equivalence checking beyond syntactic rule matching.
  \item \textbf{TM-015 governance\_deadlock:} Requires a simulator
        scenario testing whether COS can enter a state where no
        governance action is possible. Mitigation validation requires
        the sentinel escalation path (Layer 7) to be exercised.
\end{itemize}

\subsection{Public trace corpus}

v1.1 will publish a sanitized trace corpus containing:
drift events, rollback sequences, membrane activation logs, and
CMV disagreement cases drawn from the three v0.1 probes. The corpus
will be archived on Zenodo with a DOI and released under CC~BY~4.0.
It will include per-entry registry annotations linking each trace
to its source threat class, sim scenario, and harness probe.

\subsection{Expanded harness}

The v1.1 harness suite will add independent adversarial prompt
construction for HAR-SPEC-005 Class~A prompts, expanded task suites
for HAR-SPEC-012 T3, and the first probe for TM-009. The full 15-threat
coverage map is the v1.1 milestone.
